---
title: "visualization"
author: "John Quam"
date: "2023-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
library(tidyverse)
library(ggplot2)
library(bslib)
library(DT)
library(thematic)
```

## Data Processing

**These csv files I'm loading in were acquired in a previous project of mine**

**I want high percentiles to indicate a "good" measure. For metrics like Draft Age and TOV% where we want low values, (want the player to be younger and turn the ball over less) I took 1 - percentile rank already to indicate this. So being in the top percentiles of Draft Age means you're younger.**

**The percentile ranks are by position as well**

**The full data set the percentile rankings are generated from are players who played at least 75% of their team's games in college and also posted a 3rd year BPM in the NBA**

```{r}
#read in the data
percentiles = read.csv("visualization_data.csv") %>%
  #The percentiles were calculated with everyone in mind, but I'm only interested in visualizing this years draft class so I'm filtering
  filter(Draft.Year == 2022) %>%
  #rename the columns
  rename("Draft Year" = "Draft.Year", "Draft Age" = "Draft.Age", "Standing Reach" = "Standing.Reach", "RSCI Ranking" = "RSCI.Ranking",
         "TS%" = "TS.", "eFG%" = "eFG.", "ORB%" = "ORB.", "DRB%" = "DRB.", "TRB%" = "TRB.", "AST%" = "AST.", "STL%" = "STL.", 
         "BLK%" = "BLK.", "TOV%" = "TOV.", "USG%" = "USG.", "College OBPM" = "College.OBPM", "College DBPM" = "College.DBPM",
         "College BPM" = "College.BPM")

#inspect the data 
head(percentiles)
```

**Read in and modify the historical prospect data (172 historical prospects)**

```{r}
historical = read.csv("visualization_data.csv") %>%
  #filter out the current prospects for historical dataset
  filter(Draft.Year != 2022) %>%
  #rename the columns
  rename("Draft Year" = "Draft.Year", "Draft Age" = "Draft.Age", "Standing Reach" = "Standing.Reach", "RSCI Ranking" = "RSCI.Ranking",
         "TS%" = "TS.", "eFG%" = "eFG.", "ORB%" = "ORB.", "DRB%" = "DRB.", "TRB%" = "TRB.", "AST%" = "AST.", "STL%" = "STL.", 
         "BLK%" = "BLK.", "TOV%" = "TOV.", "USG%" = "USG.", "College OBPM" = "College.OBPM", "College DBPM" = "College.DBPM",
         "College BPM" = "College.BPM")
```

**Read in and modify the player stats data**

```{r}
withdrawal_players = c("Drew Timme", "Matthew Mayer", "Jalen Wilson", "Julian Strawther", "Terquavion Smith", "Harrison Ingram")

#read in the player stats
player_stats = read.csv("modeling_2022.csv") %>%
  #filter out the players that withdrew from the draft
  filter(!Player %in% withdrawal_players) %>%
  #rename the columns
  rename("Draft Year" = "Draft.Year", "Draft Age" = "Draft.Age", "Standing Reach" = "Standing.Reach", "RSCI Ranking" = "RSCI.Ranking",
         "TS%" = "TS.", "eFG%" = "eFG.", "ORB%" = "ORB.", "DRB%" = "DRB.", "TRB%" = "TRB.", "AST%" = "AST.", "STL%" = "STL.", 
         "BLK%" = "BLK.", "TOV%" = "TOV.", "USG%" = "USG.", "College OBPM" = "College.OBPM", "College DBPM" = "College.DBPM",
         "College BPM" = "College.BPM")
#inspect the data
head(player_stats)
```


### Define and Test the Function Used in the Shiny Plot

**Test the function and corresponding data frame filtering to output the plot**

```{r}
player = "Chet Holmgren"

#this is the code to filter the dataframe down to pass it to the visualization function
test = percentiles %>%
  #filter by player TODO: change this to take the ui input
  filter(Player == player) %>%
  #these columns aren't numeric so I don't want them for the plot
  select(-c("Player", "Position", "Draft Year", "Class", "School", "RSCI Ranking")) %>%
  #I want 2 columns (one with the column names and another with the corresponding percentiles) as opposed to 1 row with 19 columns
  pivot_longer(everything(), names_to = "Column", values_to = "Percentiles")

#background, text
color_hexes = c("#101010", "#FDF7F7")

percentile_plot = function(df, player, cols) {
  print(df)
  #this is the code to make the visualization function
  ggplot(df, aes(x=Percentiles, y=Column, col=Percentiles)) +
    #plot the point
    geom_point(aes(size=2)) +
    #don't want a y lable
    labs(x="Percentile Rank", y="", title = paste0("Percentile Ranks for ", player)) +
    #want the x axis scale to be from 0 to 1 (whole percentile scale)
    xlim(c(0,1)) +
    #want a color gradient with blue indicating a weakness to red indicating a strength
    scale_color_gradient(low = "lightblue", high = "red") +
    #don't need to include size in the legend as points are all the same size (just wanted them bigger)
    guides(size="none") +
    #set a black and white them
    theme(
      panel.border = element_rect(color = color_hexes[2], fill=NA),
      plot.background = element_rect(fill = color_hexes[1]),
      panel.background = element_rect(fill = color_hexes[1]),
      panel.grid.major = element_line(color = color_hexes[2]),
      panel.grid.minor = element_line(color = color_hexes[2]),
      title = element_text(color = color_hexes[2], size=24),
      axis.title.x = element_text(color = color_hexes[2]),
      axis.text.x = element_text(color = color_hexes[2]),
      axis.text.y = element_text(color = color_hexes[2]),
      axis.text = element_text(size = 16),
      legend.text = element_text(color = color_hexes[2], size=12),
      legend.title = element_text(color = color_hexes[2]),
      legend.background = element_rect(fill = color_hexes[1])
    )
}

#Test example
percentile_plot(test, player, color_hexes)
```



**Test the function and corresponding data frame filtering to output the data table**

```{r}
#function that takes the clicked value, takes the historical dataframe, finds the distance between the clicked point and 
#1. Get the x (value) and y (feature) coordinates of the clicked value 
#2. Get the position of the player selected
#3. Get the historical dataframe
#4. Create a column that takes the absolute value of the |x input - historical df value| for each player in the historical df
#5. Sort by that column to find the closest historical comps
#6. Select just the columns with player name, position, draft year, and school 
#7. Take the top 3???

get_similar_players = function(val, col, pos, df=historical) {
  #this pipeline finds 5 the most similar historical comparisons for the selected data point
  df %>% 
    #filter to only the position of the selected player (percentiles were calculated by position)
    filter(Position == pos) %>%
    #create a distance column by taking the absolute value between the player's value (x-axis click input) and the historical values
    mutate(Dist = abs(val - .data[[col]])) %>%
    #arrange to get the 5 closest values
    arrange(Dist) %>%
    #select the appropriate columns
    select(Player, Position, `Draft Year`, School) %>%
    #take the 3 closest values
    head(3)
  
}

#Test example
get_similar_players(0.74, "College OBPM", "SF") #3 according to plot should be Miles Bridges, Deandre Hunter, and Rodney Hood
```



## UI and Server Components

**Text to help with shiny app interpretation**

```{r}
#text to help viewer understand the plots
descriptive_text = "This visual allows you to select a 2022 NBA draft prospect and view what percentile rank they are in each category relative to my full NBA draft prospect dataset going back to the 2010-11 season (the first year BPM was calculated for college basketball). Percentiles were computed with respect to basketball position to account for differences between certain positions (Ex: Centers are taller than Point Guards), and high percentile ranks represent a strenght in all cases. If the 'desirable' trait corresponds to a low numerical value (Ex: all else equal, younger prospects are more desirable) I took 1 minus the percentile rank to indicate the more 'desirable' trait to be a strength. The full dataset contains 172 players, but only cosists of players who have posted a 3rd year NBA BPM (excludes all prospects from 2019-20 and 2020-21 seasons by default). The goal is to compare this year's NBA draft prospects to past prospects that stuck in the NBA as well as each other, hence the 3rd year BPM criteria. Another function of this plot is that you can click a point to see a data table with the 3 most similar players to the selected prospect for the given category in the historical dataset. For example, if I clicked the point corresponding to AJ Griffin's Wingspan percentile, I would see that his wingspan was most similar to Dorian Finney-Smith. If no point is selected, player information for all 2022 draft prospects is displayed. To return to viewing the full table after clicking a point, refresh the page. Overall, this application provides a general sense of 2022 prospects's strengths and weaknesses as well as historical comparisons for specific features. " 
#link to my github repo
github_link = "<br>If you would like more context regarding the data please check out the README file on my<a href='https://github.com/jquam15/nba_draft_visualization'> github</a>"
#link to my website
web_link = "If you'd like to see more cool content check out my <a href='https://jquam15.github.io/'>website</a>"
```


**UI and Server components**


```{r}
#get a list of unique player names
players = sort(unique(percentiles$Player))

ui <- fluidPage(
  #customize theme using bslib bs_theme() function
  theme = bs_theme(
    bootswatch = "darkly",
    base_font = font_google("PT Serif")
  ),
  #carry the theme (font style) throughout the table and plots
  thematic_shiny(font="auto"),
  #add title
  column(8, align="center", offset=2, titlePanel("2022 NBA Draft Prospect Visualization")),
  #slider input to select a player from the data (can only select 1 at a time) and style the slider accoring to css
  selectInput("player", "Player", players, multiple = F),
  #create the first row with the plot output and text description
  fluidRow(
    #set up the plot output and click elements
    column(7, align="right", plotOutput("percentiles", click="plot_click")),
    #set up the descriptive text and web link elements
    column(5, align="left", htmlOutput("descriptive_text"),#, style="color:#BFA181"), 
           htmlOutput("github_link"),#, style="color:#BFA181"), 
           htmlOutput("web_link")#, style="color:#BFA181")
           )
  ),
  #add the second row with the data table
  div(
    #want to add some space between the table and the text/plot output
    style = "margin-top: 3em;",
    #make a row with a data table output
    fluidRow(
      column(12, DT::dataTableOutput("table"))
    )
  ),
)

server <- function(input, output) {
  #render the plot
  output$percentiles = renderPlot({
    percentiles %>%
      #filter by player 
      filter(Player %in% input$player) %>%
      #these columns aren't numeric so I don't want them for the plot
      select(-c("Player", "Position", "Draft Year", "Class", "School", "RSCI Ranking")) %>%
      #I want 2 columns (one with the column names and another with the corresponding percentiles) as opposed to 1 row with 19 columns
      pivot_longer(everything(), names_to = "Column", values_to = "Percentiles") %>%
      percentile_plot(input$player, color_hexes) 
  }, bg="transparent")
  
  
  #render a data table which by default will be this year's class basic info, but will change to a historical comp upon a user click
  output$table = DT::renderDataTable({
    #if there is no click yet display the current class's player information as the data table
    if (is.null(input$plot_click)) {
      output_df = player_stats %>%
        #only take specific columns
        select(c("Player", "Position", "Draft Age", "Class", "School", "Height", "Weight", "RSCI Ranking", "College OBPM",
                 "College DBPM", "College BPM"))
    #if there is a click then find the 3 most similar players at that position to the selected player in the selected category and return that as the table
    } else {
      #get all column names
      columns = colnames(percentiles)
      #remove columns not in plot
      #order columns so that the column you click sends a y value that can be used to index this vector so that the vector[y_input] = desired column
      columns = sort(columns[!columns %in% c("Player", "Position", "Draft Year", "Class", "School", "RSCI Ranking")])
      #get the column the user selected and store it so it can be passed to function (need to round to integer) that gets similar historical comps
      col = columns[round(input$plot_click$y)]
      #get x-axis user input (percentile value) and store it in variable to be passed to function that gets similar historical comps
      #round for a little more accuracy
      val = round(input$plot_click$x, 2)
      #get the player's position as percentiles were calculated according to position
      pos = percentiles[percentiles$Player == input$player, "Position"]
      #call the function with the appropriate information designated by the user input
      output_df = get_similar_players(val, col, pos)
    }
    #display the table according to whether user input has been fully given or not
    output_df
  })
  
  
  #this renders the first paragraph of descriptive text
  output$descriptive_text = renderText({
    paste("", descriptive_text, sep="\t")
  })

  #this provides a link to the code
  output$github_link = renderText({
    paste(github_link)
  })
  
  #this provides a link to my website homepage
  output$web_link = renderText({
    paste0("<br>", web_link)
  })
}
```


## Shiny App

```{r}
shinyApp(ui, server)
```





